---
title: "death_rate_prep"
author: "Edelweiss Gevorgyan"
date: "`r Sys.Date()`"
output: pdf_document
---


```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(cowplot)
library(patchwork)
library(dplyr)
library(ggrepel)
library(tidyverse)
library(gridExtra)
library(grid)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```



Preparation for Age_Gender_2023 Data

```{r age_gender_2023}
age_gender <- read_excel("../Data/age-gender.xlsx", skip = 4)





colnames(age_gender) <- c("Age", "Male_2011", "Female_2011", "Total_2011", 
                          "Male_2012", "Female_2012", "Total_2012", 
                          "Male_2013", "Female_2013", "Total_2013", 
                          "Male_2014", "Female_2014", "Total_2014", 
                          "Male_2015", "Female_2015", "Total_2015", 
                          "Male_2016", "Female_2016", "Total_2016", 
                          "Male_2017", "Female_2017", "Total_2017", 
                          "Male_2018", "Female_2018", "Total_2018", 
                          "Male_2019", "Female_2019", "Total_2019", 
                          "Male_2020", "Female_2020", "Total_2020", 
                          "Male_2021", "Female_2021", "Total_2021", 
                          "Male_2022", "Female_2022", "Total_2022", 
                          "Male_2023", "Female_2023", "Total_2023")

age_gender_2023 <- age_gender %>%
  slice(-1) %>%
  select(Age, Male_2023, Female_2023, Total_2023)

head(age_gender_2023)
```

```{r}
getwd()
```

Armenia death rate data
```{r death_data}
data <- read.csv("../Data/death_Armenia.csv", skip = 8, stringsAsFactors = FALSE)
data$Year <- as.numeric(substr(data$Date, 1, 4))
data <- data[data$Year <= 2024, ]
data <- data[ , !names(data) %in% "Annual...Change"]
names(data) <- c("Date", "Death Rate", "Annual Change", "Year")

death_data <- data
head(death_data)
```


Death Rate Data for Azerbaijan 

```{r death_data_az}
data_azerbaijan <- read.csv("../Data/death_Azerbaijan.csv", skip = 8, stringsAsFactors = FALSE)

data_azerbaijan$Year <- as.numeric(substr(data_azerbaijan$Date, 1, 4))

data_azerbaijan <- data_azerbaijan[data_azerbaijan$Year <= 2024, ]

data_azerbaijan <- data_azerbaijan[ , !names(data_azerbaijan) %in% "Annual...Change"]

names(data_azerbaijan) <- c("Date", "Death Rate", "Annual Change", "Year")

death_data_az <- data_azerbaijan
```

Death Rate Data for Georgia 

```{r death_data_ge}
data_georgia <- read.csv("../Data/death_Georgia.csv", skip = 8, stringsAsFactors = FALSE)
data_georgia$Year <- as.numeric(substr(data_georgia$Date, 1, 4))
data_georgia <- data_georgia[data_georgia$Year <= 2024, ]
data_georgia <- data_georgia[ , !names(data_georgia) %in% "Annual...Change"]
names(data_georgia) <- c("Date", "Death Rate", "Annual Change", "Year")
death_data_ge <- data_georgia
```

Heathcare Spending in ARmenia in USD 

```{r data_armenia_healthcare}
data_armenia_healthcare <- read.csv("../Data/armenia-healthcare-spending.csv", skip = 8, stringsAsFactors = FALSE)
data_armenia_healthcare$Year <- as.numeric(substr(data_armenia_healthcare$Date, 1, 4))
data_armenia_healthcare <- data_armenia_healthcare[data_armenia_healthcare$Year <= 2024, ]
data_armenia_healthcare <- data_armenia_healthcare[ , !names(data_armenia_healthcare) %in% "Annual Change"]
names(data_armenia_healthcare) <- c("Date", "Per Capita (US $)", "% of GDP", "Year")
data_armenia_healthcare$Year <- as.numeric(data_armenia_healthcare$Year)
data_armenia_healthcare$`Per Capita (US $)` <- as.numeric(data_armenia_healthcare$`Per Capita (US $)`)
data_armenia_healthcare$`% of GDP` <- as.numeric(data_armenia_healthcare$`% of GDP`)
data_armenia_healthcare <- data_armenia_healthcare[, c("Date", "Year", "Per Capita (US $)", "% of GDP")]
data_armenia_healthcare$Year <- format(as.Date(data_armenia_healthcare$Date), "%Y")
data_armenia_healthcare$Year <- as.numeric(data_armenia_healthcare$Year)

```



Infant Mortality Rate 

```{r infant_mortality_data}
data_infant_mortality <- read.csv("../Data/infant_mortality.csv", skip = 8, stringsAsFactors = FALSE)
data_infant_mortality$Date <- as.Date(data_infant_mortality$Date, format = "%Y-%m-%d")
data_infant_mortality$Year <- as.numeric(format(data_infant_mortality$Date, "%Y"))
data_infant_mortality <- data_infant_mortality[data_infant_mortality$Year <= 2024, ]
data_infant_mortality <- data_infant_mortality[, !names(data_infant_mortality) %in% "Annual % Change"]
names(data_infant_mortality) <- c("Date", "Death per 1000 Live Births", "Annual % Change", "Year")
data_infant_mortality$`Death per 1000 Live Births` <- as.numeric(data_infant_mortality$`Death per 1000 Live Births`)
data_infant_mortality$`Annual % Change` <- as.numeric(data_infant_mortality$`Annual % Change`)
data_infant_mortality <- data_infant_mortality[, c("Date", "Year", "Death per 1000 Live Births", "Annual % Change")]
data_infant_mortality$Year <- format(as.Date(data_infant_mortality$Date), "%Y")
data_infant_mortality$Year <- as.numeric(data_infant_mortality$Year)
data_infant_mortality <- data_infant_mortality[data_infant_mortality$Year <= 2024, ]
infant_mortality_data <- data_infant_mortality
```



Data for Pyramid 2023 for AGe and Gender Distribution
```{r age_gender_data_2023}
age_gender_data_2023 <- age_gender_2023 %>%
  filter(Age != "Total") %>%
  mutate(Age = suppressWarnings(as.numeric(Age))) %>%
  filter(!is.na(Age)) %>%
  mutate(AgeGroup = cut(Age, 
                        breaks = seq(0, 100, by = 5), 
                        right = FALSE, 
                        labels = paste(seq(0, 95, by = 5), seq(4, 99, by = 5), sep = "-"))) %>%
  group_by(AgeGroup) %>%
  summarise(Male = sum(Male_2023, na.rm = TRUE), 
            Female = sum(Female_2023, na.rm = TRUE), 
            .groups = 'drop') %>%
  pivot_longer(cols = c(Male, Female), names_to = "Gender", values_to = "Population") %>%
  mutate(Population = ifelse(Gender == "Male", -Population, Population))

```

Age-Gender Pyramid Plot

```{r}
pyramid_2023 <- ggplot(age_gender_data_2023, aes(x = AgeGroup, y = Population, fill = Gender)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = abs(Population)), position = position_stack(vjust = 0.5), size = 3) +
  coord_flip() +
  scale_y_continuous(labels = abs) +
  scale_fill_manual(values = c("Male" = "lightblue", "Female" = "pink")) +
  labs(title = "Population Pyramid of Armenia (2023)",
       x = "Age Group",
       y = "Population") +
  theme_classic() +
  theme(legend.title = element_blank())

```


Line Graph for Death Rate in Armenia

```{r death_rate_plot}
death_rate_plot <- ggplot(death_data, aes(x = Year, y = `Death Rate`)) +
  geom_line() +
  labs(title = "Death Rate in Armenia Over Time", 
       x = "Year", 
       y = "Death Rate") +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        legend.position = "none")


death_annual_change_plot <- ggplot(death_data, aes(x = Year, y = `Annual Change`)) +
  geom_line() +
  labs(title = "Annual Change in Death Rate in Armenia Over Time", 
       x = "Year", 
       y = "Annual Change (%)") +
  theme_minimal()

grid.arrange(death_rate_plot, death_annual_change_plot, ncol = 1)
```

```{r}
colnames(death_data)
```

```{r}
death_rate_plot_events <- ggplot(death_data, aes(x = Year, y = `Death Rate`)) +
  geom_line() +
  geom_vline(xintercept = c(1950, 1973), color = "green", linetype = "dashed") + 
  
  geom_point(aes(color = "Widespread Vaccination"),
             data = death_data %>% filter(Year == 1960), 
             size = 3) +
  geom_vline(xintercept = c(1975, 1995), color = "red", linetype = "dashed") + 

  geom_point(aes(color = "Spitak Earthquake"),
             data = death_data %>% filter(Year == 1988), 
             size = 3) +  # Spitak Earthquake
  geom_point(aes(color = "1991, Collapse of USSR"),
             data = death_data %>% filter(Year == 1991), 
             size = 3) +  # Collapse of the Soviet Union
  geom_point(aes(color = "1994, End of Nagorno-Karabakh War"),
             data = death_data %>% filter(Year == 1994), 
             size = 3) +  # Artsakh War
  geom_vline(xintercept = c(2003, 2008), color = "orange", linetype = "dashed") + 

  geom_point(aes(color = "Economic Crisis"),
             data = death_data %>% filter(Year == 2008), 
             size = 3) +



  labs(title = "Death Rate in Armenia Over Time", 
       x = "Year", 
       y = "Death Rate") +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        legend.position = "bottom",
        legend.title =  element_blank(),
        legend.text = element_text(size = 6))

death_rate_plot_events
```

Combining data for comp. graph

```{r data_combined_south_caucasus}
death_data$Country <- "Armenia"
data_azerbaijan$Country <- "Azerbaijan"
data_georgia$Country <- "Georgia"

data_combined_south_caucasus <- bind_rows(death_data, data_azerbaijan, data_georgia)
colnames(data_combined_south_caucasus)
```
Plot for comparing Arm, Ge, Az death rates

```{r death_rate_caucasus_plot}
death_rate_caucasus_plot <- ggplot(data_combined_south_caucasus, aes(x = Year, y = `Death Rate`, color = Country)) +
  geom_line() +
  labs(title = "Death Rates in South Caucasus (Armenia, Azerbaijan, Georgia)", 
       x = "Year", 
       y = "Death Rate") +
  theme_minimal()
```



Data for Death causes in Armenia
```{r death_causes_data}
data_causes <- read.csv("../Data/death_causes.csv", stringsAsFactors = FALSE)

data_causes$Year <- as.numeric(data_causes$Year)
data_causes$Value <- as.numeric(data_causes$Value)

death_causes_data <- data_causes %>%
  filter(!is.na(Year), !is.na(Value)) %>%
  select(Year, Category, Value)
```

Stacked Bar Chart of Death Causes in Armenia
```{r death_causes_plot}
death_causes_plot <- ggplot(data_causes, aes(x = Year, y = Value, fill = Category)) +
  geom_bar(stat = "identity", position = "stack") +
    scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Causes of Death in Armenia Over Time",
    x = "Year",
    y = "Deaths per 100,000 Population",
    fill = "Cause"
  ) +
  theme_minimal()
```

```{r}
head(death_data)
```
```{r}
head(data_armenia_healthcare)
```


Merging data for Healthcare VS Death Rate
```{r merged_data_hd}
merged_data_hd <- merge(death_data, data_armenia_healthcare, by = "Year")
colnames(merged_data_hd) <- c("Year", "Date_death", "Death_Rate", "Annual_Change", "Date_healthcare", "Date_health", "Per_Capita_USD", "GDP_Percentage")
head(merged_data_hd)
```


Plot for Healthcare Spending VS Death Rate

```{r health_death_plot}
health_death_plot <- ggplot(merged_data_hd, aes(x = Per_Capita_USD, y = Death_Rate, color = Year)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  xlab("Healthcare Spending Per Capita (USD)") +
  ylab("Death Rate") +
  ggtitle("Relationship Between Healthcare Spending and Death Rate") +
  scale_color_gradient(low = "green", high = "black") +
  theme_minimal()
```

```{r}
health_death_plot
```

Infant Mortality Rate Over Time and Annual Change

```{r IMR_plot}
plot_IMR_change <- ggplot(infant_mortality_data, aes(x = Year, y = `Death per 1000 Live Births`)) +
  geom_line() +
  ggtitle("Infant Mortality Rate over Time") +
  xlab("Year") +
  ylab("IMR (Deaths per 1000 Live Births)") +
  theme_minimal() +
  theme(axis.title.x = element_blank())


plot_IMR_pc <- ggplot(infant_mortality_data, aes(x = Year, y = `Annual % Change`)) +
  geom_line() +
  ggtitle("Annual % Change in IMR") +
  xlab("Year") +
  ylab("Annual % Change") +
  theme_minimal()

#grid.arrange(plot_IMR_change, plot_IMR_pc, ncol = 1)
```






```{r}
save(
  pyramid_2023, death_rate_plot,
  death_annual_change_plot, death_rate_caucasus_plot,
  death_causes_plot, health_death_plot,
  plot_IMR_change, plot_IMR_pc,
  death_rate_plot_events, age_gender_data_2023,
  
  file = "../Rdata_saved/death_rate_analysis.RData"
)
```

















