---
title: "Scratch"
author: "Levon Titanyan"
date: "`r Sys.Date()`"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

```{r libraries}
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(cowplot)
library(patchwork)
library(dplyr)
library(ggrepel)
library(tidyverse)
library(gridExtra)
library(grid)
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(cowplot)
library(patchwork)
library(dplyr)
library(ggrepel)
library(tidyverse)
library(gridExtra)
library(grid)

options(scipen = 999)
load("../Rdata_saved/migration_analysis.RData")
load("../Rdata_saved/birth_rate_analysis.RData")
load("../Rdata_saved/death_rate_analysis.RData")
```

```{r}
population_histogram_by_ages

```
The graph shows the population trends in Armenia from 2011 to 2023 by four age groups: 0-18, 19-35, 36-55, and 56+. 
The 0-18 and 19-35 groups remained stable, while the 36-55 group experienced slight declines, reflecting a decrease in young adults and middle-aged individuals. The 56+ group, on the other hand, saw a steady increase, highlighting the aging population in Armenia. Overall, the graph demonstrates a trend of population aging with declines in younger age groups over the 12-year period.



```{r pyramid_2023}
pyramid_2023
```
The population pyramid graph represents the age and gender distribution of Armenia's 2023 population. The visualization uses a back-toback bar chart to compare the make and female population across different age groups. 
We can observe that the mid-age groups (30-50) have the highest population density overall. This indicates that Armenia has a large working-age population. There is relatively large population in younger age groups (0-20) as well. The population decreases notably in the older age groups, indicating a relatively smaller elderly population. 
We also observe that among younger age groups (0-20), males consistently outnumber females, which can be because of the gender-selective births. STarting from the young-mid group, females begin to outnumber males, and this trend becomes increasingly pronounced in the older age groups. This shift can be attributed to several factors, such as higher male mortality rate, natural advantage of female longevity, or migration patterns.




Here's the overall trend of birth rate with percentage change in Armenia from 1950 to 2023.
```{r birth_rate_plot}
grid.arrange(birth_rate_plot,change_percent_plot, ncol = 1)

```

```{r arm_az_geo_plot}
arm_az_geo_plot
```
This graph presents the birth rates in the Southern Caucasus region, encompassing Armenia, Georgia, and Azerbaijan, from 1950 to 2023. It highlights the similar patterns observed across the three countries, with Azerbaijan consistently exhibiting higher birth rates compared to Armenia.



As we can see during the postwar period, birth rates in Armenia and other Soviet republics were increasing as part of a broader effort to stabilize and grow the population. We hypothesize that this trend was further accelerated by pro-natalist policies introduced during this era. Historical records and research on Soviet family policies, such as those detailed in Family in the Soviet Union and reports by the Economic and Social Research Council, highlight how measures like the restoration of maternity leave in 1955 played a crucial role. Maternity leave was extended to 56 days before and up to 70 days after childbirth, with benefits covering two-thirds of wages and offered to a broader range of families. Additionally, poorer households received lump-sum support for newborn care. These policies were part of the Soviet strategy to promote demographic recovery and population growth following the war, reflecting the state’s commitment to supporting families and addressing postwar challenges.
```{r birth_rate_plot_1}
# Adding 1950-1960 years
grid.arrange(birth_rate_plot_1, change_percent_plot_1,ncol = 1, heights = c(1, 1))  
```
The data confirms that birth rates increased significantly during the postwar era, reflecting the success of pro-natalist policies in stabilizing and boosting the population. However, we observe that the 1955 policy changes, while beneficial for families, did not lead to any notable shifts in the existing trend. Instead, the steady growth in birth rates suggests that the postwar demographic recovery was primarily influenced by broader social and economic recovery efforts rather than individual policy adjustments. Thus, while policies like extended maternity leave and financial support were impactful, they complemented rather than revolutionized the ongoing birth rate growth during this period.




In the early 1960s, divorce procedures were liberalized, making it easier for couples to dissolve marriages. By the mid-1960s, the Soviet Union faced a demographic crisis, with fertility rates dropping below replacement levels. However, in 1968, restrictions on divorce were implemented, which we believe were aimed at stabilizing family structures and reversing the demographic trend. Alongside these restrictions, expanded social benefits for families with disabled children were introduced, highlighting the Soviet Union’s effort to address the demographic crisis and support vulnerable groups.We hypothesize that the restriction on divorce had a significant impact on the birth rate stabilization. 
```{r birth_rate_plot_2}
# Adding 1960-1969 years
grid.arrange(
  birth_rate_plot_2,
  change_percent_plot_2,
  ncol = 1,
  heights = c(1, 1))  
```
With the introduction of restrictions on divorce in 1968, we can see a stabilization in the birth rate, supporting the hypothesis that these policies contributed to slowing down the decline in fertility rates. Despite this stabilization, the birth rate did not rise significantly, indicating that while the divorce restrictions had some mitigating effect, they were not enough to fully reverse the broader demographic trends.






The data, taken from 'USSR and other word countries', shows the trends in marriage and divorce rates in the USSR, USA, and Japan from the 1950s to the 1980s. These countries experienced similar economic and health-related challenges, yet their responses and the effects on marriage and divorce rates differed.
The graph highlights that, despite the 1968 restrictions on divorce in the USSR, there was little noticeable impact on divorce rates. However, marriage rates in the USSR began to show a gradual increase following the 1968 law, signaling a shift in family dynamics during this period. This suggests that while the divorce restriction law had limited effect on reducing divorce rates, it may have influenced an uptick in marriages during the late 1960s and early 1970s.
```{r marriages_data_plot}
# Plot of Marriages and Divorces rate in USSR, USA and Japan
marriages_divorces_plot
```


1969: The Code of Laws on Marriage and Family Life was introduced, aiming to promote gender equality within families, improve family stability, and provide greater protection for vulnerable spouses, marking a shift toward prioritizing family welfare.

In 1974, the Soviet Union implemented new family support policies to address declining birth rates and support vulnerable populations. Monthly child allowances were introduced for families with per capita incomes below 50 rubles (75 rubles in the Extreme North and Far East regions), providing 12 rubles per child under the age of 8. These measures aimed to alleviate poverty and stimulate fertility, though their primary impact was on improving the living conditions of lower-income families.

The momentum continued in 1975 when the United Nations designated it as International Women’s Year, prompting a global focus on ending discrimination against women. The USSR responded by emphasizing the role of women in society and further advocating for policies that improved conditions for women balancing work, motherhood, and household responsibilities. These combined efforts reflected a broader shift towards fostering gender equality and stabilizing demographic trends.




During the 1960s and 1970s, a series of social reforms were implemented in the Soviet Union to address declining birth rates and promote family welfare. The 1969 Code of Laws on Marriage and Family Life, which aimed to promote gender equality and family stability, coincided with the early decline in fertility rates. Additionally, the 1974 family support policies, which provided financial support to lower-income families, aimed to incentivize childbirth by improving living conditions. Based on these policies, we hypothesize that these social reforms had a significant impact on stabilizing birth rates and improving family welfare in Armenia.
```{r birth_rate_plot_3}
# adding 1969-1987 years
grid.arrange(
  birth_rate_plot_3,
  change_percent_plot_3,
  ncol = 1,
  heights = c(1, 1))
```
We can conclude that the introduction of the 1969 marriage equality reforms had a limited effect on birth rates in Armenia, with little noticeable change in the demographic trends. However, following the implementation of the 1974 family support policies aimed at assisting low-income families, birth rates began to stabilize and even showed a slight increase. This marked the first time in 20 years that birth rates had increased, signaling a positive shift in demographic trends due to the social reforms introduced during this period.
It is also worth mentioning that 1975 was designated as International Women’s Year by the United Nations, a global movement that emphasized gender equality. During this period, the USSR also focused on improving conditions for women balancing work, motherhood, and household responsibilities, contributing to a broader shift toward gender equality and supporting demographic stability.




The period between 1985 and 2000 was one of profound change for Armenia, shaped by a series of significant events that had lasting impacts on the country’s socio-political and economic landscape. These events disrupted daily life, strained resources, and triggered large-scale migration. Given the severity of these disruptions, we hypothesize that these events had a negative impact on Armenia’s birth rate during this period, potentially leading to a decline in birth rates and shifts in family structures as a result of economic hardship, displacement, and social instability.
```{r birth_rate_plot_of_90s}
# Plot of Birth rate in 90's (1985-2000 is shown for broader range and better understanding)
plot_of_90s
```
The 1988 Spitak Earthquake, followed by the Artsakh War from 1988 to 1994, and the  collapse of the Soviet Union during the war (1991 year), led to a significant decline in Armenia’s birth rate. From 1985, the birth rate dropped from 24.1 to 13.5 by 2000, reflecting the socio-economic disruptions caused by these events. The initial impact of the earthquake and the ongoing war resulted in a sharp decline in the birth rate. Following the end of the Artsakh War, the birth rate began to stabilize, showing signs of recovery in the late 1990s, though it remained below the pre-crisis levels. This trend highlights the profound influence of these major events on Armenia’s demographic patterns during this period.


```{r birth_rate_plot_4}
# Combine the plots for 1988-2000
grid.arrange(
  birth_rate_plot_4,
  change_percent_plot_4,
  ncol = 1,
  heights = c(1, 1))
```




During the 2008 global economic crisis, financial instability and uncertainty were felt across many countries, and it is hypothesized that these factors significantly influenced birth rates. The crisis led to reduced economic opportunities, job insecurity, and financial stress for families, likely resulting in delayed or reduced childbearing. Similarly, the COVID-19 pandemic in 2020, with its health crises, lockdowns, and economic disruptions, is expected to have had a major effect on birth rates. This would have been driven by increased uncertainty about the future, job losses, and shifts in social and healthcare priorities.
In this context, we compare Armenia with Germany, France, and Italy to analyze the varying impacts of these global events on birth rates in European countries 

```{r comparison_plot_2000_2023}
comparison_plot_2000_2023
```
The analysis indicates that while the 2008 economic crisis had a significant and immediate impact on birth rates, leading to a sharp slow down in growth in Armenia, France and Italy, whereas in Germany it had opposite impact: While having negative trend in birth rate, after 2008, it stabilized.

the COVID-19 pandemic of 2020 had  no effect on birth rates in 4 countries in this short-term data. This suggests that economic crises tend to have a more immediate and noticeable influence on demographic trends, while health-related crises like the pandemic may affect birth rates in more complex ways or over a longer duration.

```{r birth_rate_plot_5}
grid.arrange(
  birth_rate_plot_5,
  change_percent_plot_5,
  ncol = 1,
  heights = c(1, 1)
)
```




# death






```{r death_rate_plot, warning=FALSE}
grid.arrange(death_rate_plot, death_annual_change_plot, ncol = 1)
```
The graph illustrates the changes in Armenia's death rate over time, spanning the period from 1950 to 2024. Between 1950 and 1975, the death rate shows a steady decline, which likely reflects the significant imporvements in public health infrastructure, continous advances in medical care and enhanced living conditions. Between 1975 and 1990, the death rate stays relatively stable. In 1990, there is a clear increase in death rate, mainly because of the First Artsakh War. Starting from here onwards, the decreasing trend shifts, with the death rate beginning to increase. 



```{r}
death_rate_plot_events
```
From 1950 to 1975, Armenia experienced a steady decline in its death rate, primarily due to advancements in healthcare and significant public health measures during the Soviet era. A cornerstone of this progress was the widespread implementation of vaccination campaigns, which drastically reduced mortality from infectious diseases such as polio, tuberculosis, and measles. These campaigns, supported by centralized Soviet healthcare policies, ensured high immunization coverage across the population, creating robust herd immunity. Alongside vaccinations, improvements in access to medical services, sanitation, and nutrition further enhanced public health outcomes. Economic stability during this period also played a role in supporting these initiatives, leading to a consistent reduction in death rates and increased life expectancy.


Between 1975 and 2000, Armenia's death rate began to rise, driven by a combination of socio-political and economic challenges. The late 1970s marked the start of stagnation in the Soviet Union, leading to reduced investment in healthcare infrastructure and a decline in public health services. This was compounded by the devastating Spitak earthquake in 1988, which claimed over 25,000 lives and severely strained the country’s healthcare system. Following the collapse of the Soviet Union in 1991, Armenia faced a period of economic turmoil, with hyperinflation, widespread poverty, and a breakdown of social services. The protracted Nagorno-Karabakh conflict (1988–1994) further exacerbated the situation, contributing to casualties, displacement, and limited access to medical care. During this time, rising mortality from chronic conditions such as cardiovascular diseases, worsened by inadequate healthcare resources and deteriorating living conditions, became a significant factor in the increasing death rate.









The collapse of the Soviet Union in 1991 led to a significant increase in death rates in Armenia due to economic instability, disruptions in healthcare systems, and declining living conditions. 
```{r}
death_rate_caucasus_plot
```
The graph compares death rates across the South Caucasus countries - Armenia, Georgia and Azerbaijan. We can observe that initially Armenia had a lower death rate compared to that of its neighbouring countries, highlighting a better demographic conditions. In 1970, Armenia achieved the lowest death rates in the region, while the other two countries still remained higher. Despite the increase in death rates started from 1980s, Armenia's death rate remains consistently lower. While Georgia's rate peaked significantly after this period, Armenia maintained a steady trajectory, avoiding sharp increases. 






Given the high prevalence of risk factors such as smoking, alcohol consumption, poor dietary habits, and physical inactivity, coupled with limited healthcare access and preventive measures, Armenia has seen a significant rise in cardiovascular disease (CVD)-related deaths over time, particularly in the post-Soviet era, where systemic healthcare challenges and lifestyle factors have exacerbated the burden of CVD on the population.

```{r}
death_causes_plot
```
The bar chart illustrates the primary causes of death in Armenia from 2016 to 2022, measured per 100,000 people. Cardiovascular diseases are the leading cause of death throughout the observed period, consistently accounting for the largest portion of fatalities. Cancer ranks second, contributing significantly but at a lower rate than cardiovascular diseases. Chronic respiratory diseases and diabetes are comparatively minor causes, with both remaining stable and contributing only a small fraction of the total deaths.





Hypothesis: Higher healthcare spending per capita leads to decrease in the death rate, as better funding should improve healthcare outcomes.

```{r health_death_plot}
health_death_plot
```
The scatter examines the relationship between healthcare spending per capita and death rates in Armenia over time, with data points being shaded by years. The observed positive correlation presents a counterintuitive trend, as higher healthcare expenditures are usually associated by lower death rates. This could be explained by several factors. First, Armenia's aging population may lead to higher healthcare needs and mortality rates. Also, the rising costs of healthcare, including inflation and expensive medical technologies contribute to increased spending without improving health condition.


```{r}
grid.arrange(plot_IMR_change, plot_IMR_pc, ncol = 1)

```
The graph reveals a clear long-term decline in infant mortality in Armenia, showcasing significant progress in healthcare and living standards over the decades. From 1950 to 1990, the steady reduction in IMR aligns with systematic investments in public health under the Soviet Union, including widespread access to maternal and child healthcare.
Economic hardships and political instability during the early post-Soviet era likely disrupted healthcare delivery, slowing progress temporarily. The stabilization and renewed decline in IMR from the 2000s onward reflect the recovery of the healthcare system, supported by economic reforms, international aid, and improved governance.




# migration



Migration has been a defining feature of Armenia’s history, influenced by its geopolitical position, economic conditions, and socio-political developments. The patterns of migration in Armenia are deeply intertwined with its history, and are uncovering the changes the country has experienced over time. 

In this section we are exploring the historical development of migration in Armenia, and uncovering the underlying patterns and reasons for them. The aim of the research is to identify the main factors affecting migration in Armenia. The findings will be useful in the sense of predicting and controlling the future inflows and outflows of population. 

With the research we want to check the hypothesis if improvements in overall quality of life in Armenia are positively correlated with net migration, as individuals are more likely to move to areas with better living standards and opportunities.

Starting with the historical development of migration of Armenia, it is important to have a glance on the overall pattern for the previous 60 years. This uncovers the main direction the country has been moving and gives a general understanding of the main events that have been affecting the migration.

```{r}
migration_plot
```

The line chart above summarizes net migration in Armenia starting from 1960 till 2023. It is clearly divided into three main sections: Soviet times, transitional period and independence era. Having a positive net migration until 1978 uncovers that Soviet times have been favorable for life in Armenia and immigration into the country has been bigger than emigration. One of the reasons for that might be the inflow of Armenians from Syria and Lebanon in the 60s. The metric became negative around 1978 when Armenians starting immigrating to the USA as Soviet borders became more permissive. 

The period of transition to an independent country has shaken all the spheres of the country and migration is not an exception. The "shake" caused by the political situation is evident from the graph. With the start of Independence movement in 1988, massive inflow of population to fatherland resulted in high value of net migration. In just 3 years the emigration from the country has reached its historical low of 200,000 as a result of a extremely tough economic situation and a severe war going on in Artsakh. After the end of the war, net migration has increased since less people where leaving the country as well as a lot of people relocated to Armenia fro Artsakh. But still, the net migration value has never become positive, meaning that during the whole period of independent Armenia more people are leaving the country than entering it.

The stabilization of the political and economical situation after the year 2000, has lead to a more stable value of net migration over the time. The start of COVID-19 in 2020 has had large impact on emigration and has decreased it significantly, mainly because the people could not leave the country. Moreover, as a result of Artsakh Second War, many people from Artsakh emigrated to Armenia, increasing the net migration. Another rapid increase in net migration is observed in 2022, when the Russian-Ukrainian conflict has started. Due to very large immigration from Russia to Armenia, the net migration has increased, but still remained negative, meaning that emigration has been still larger.

Overall, the graph highlights how major events in the political and economical areas shake the migration image and affect the population change.

To analyze the net migration for the recent years it is important to look onto the yearly border crossings of Armenia. While it does not give an exact picture of migration, it uncovers trends that might lead to migation changes in the future.

```{r}
border_plot
```

The chart summarizes the number of arrivals and departures in Armenia. The net column which shows the difference between arrivals and departure uncovers that while in 2019 more people where departing from Armenia than entering Armenia, the picture stated to change from 2020, when the net value became positive yet very small. Still, while this is not a very accurate way to indicate a positive flow into the country, it summarizes that there is no ubrupt inflow or outflow in the recent years, and gives a picture of stability.

To identify any major differences between Armenian and Non-Armenian citizens entering and leaving Armenia we have looked into the arrivals and departures each year for both Armenian and Foreign citizens.

```{r}
border_arm_for_plot
```

The graph shows that in ther period 2016-2021, the proportion of arrivals and departures have been approximately the same for both Armenian and foreign citizens. This means that there is no year where Armenian citizens massively left the country making the proportion of departures higher and there is no year when foreign citizens massively entered the country making the proportion of arrivals higher. So, the picture is more or less stable. Still, the graph shows that Armenian citizens cross the borders more often than foreign citizen which indicates that movements of Armenians across the borders dominate in the country and therefore, Armenian citizens affect the migration statistics more than foreigners. 

Each time people enter a new country or leave it, they are driven by some set of reasons. As the goal of the research is to identify the key factors affecting population change and in this case migration, it is important to be familiar with the reasons that drive people across countries.

For this purpose, we have looked into the main reason people immigrated to Armenia and emigrated out of Armenia in 2017.

```{r}
grid.arrange(reasons_imm, reasons_em, heights = c(1.35, 1))
```

The main reason driving people to move to Armenia or leave from Armenia is seasonal work. On the second place in both cases is non-seasonal work, up to 1 year. This result shows that in general work opportunities are what drivers people to change their country of residence. The main insight the charts provide is that a large proportion of people leave from Armenia for permanent residence while very few come to Armenia for that reason. And our goal is to find the reason behind this and identify the factors that influence people to leave Armenia permanently.


Analizing factors one by one will give a comprehensive image on what is the key factor that drives the population out of the country. We will focus on the factors the literature suggests have the most impact and will test each of them to see how applicable are the literature-suggested factors in case of Armenia.

We have defines three main types of factors that might affect migration: economic, general well-being and social

As suggested by initial graph on overall historical migration in Armenia, the economy has a high influence on migration levels. As the main economic indicator we have chosen GDP per Capita calculated based on People Purchasing Power. Another economic factor we take into account is the unemployment rate, since as shown previously, work is the main reason people decide to relocate. With the analysis, we aim to test the hypothesis that higher GDP per Capita  and lower unemployment rate result in higher net migration, as a result of making the country more favourable place to live.

```{r}
grid.arrange(gdp_plot, unemp_plot)
```

As the graphs suggest, there is a positive correlation between GDP per Capita and net migration, while there is also not very strong but negative correlation between unemployment rate and net migration. Therefore, the hypothesis can be accepted.


The literature defines that general well-being is also an important factor, which is why we are also looking into the correlation between happiness index and net migration. We want to test if higher happiness index means bigger net migration.

```{r}
happiness_plot
```

While the data points are too scarce to make strong conclusions, still there is some correlation, so there is no evidence to reject the null hypothesis. Therefore, we can claim that there is a positive correlation between happiness index and net migration to some extent. This means that the happiness in a country (specifically Armenia) supports the decision on not to leave the country.


Coming to the social reasons, it is important to mention medical care. To analyze the factor we looked at the correlation of health expenditure as a percentage of GDP and net migration. Another important social factor is education, therefore education expenditure as a percentage of GDP vs net migration is also explored. So, the nex hypothesis is if higher expenditures on social aspects like healthcare and education have positive correlation with net migration.

```{r}
grid.arrange(health_plot, edu_plot)
```

While in case of health expenditure the correlation is evident, education expenditure does not show any valuable results, since all the points are gathered in the same area and do not show any direction. Therefore while health expenditure can be considered an influencing factor on net migration, education has no connection to that.

So, the analysis identifies that economic factors are the ones that have the most significant influence on migration in Armenia.The correlation can be seen in almost all of the cases, so the initial hypothesis on higher qualitiy of life results in increased migration is proved. While most of the factors show some kind of correlation and might actually have their impact on migration, there is an important aspect that should not be neglected. If we look at the time colors of the points we can see that most of the factors increase over time (except unemployment rate which decreases over time). We also saw that net migration also increases over time in recent years. Therefore, it is hard to define if the two factors are actually connected or both are increasing as the time goes as a result of country's development. We still believe that there is some significant correlation but we cannot conclude anything about causation.








